---
title: "MATH5845 --- Time Series Analysis: Seasonal ARIMA model for West Virginia Beer Sales"
author:
  - Department of Statistics
  - University of New South Wales
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---
  
  
# Packages
```{r, message=FALSE}
library(astsa)
#data(package="astsa")
library(itsmr)
```



```{r, message=FALSE}

Data<-read.csv("WEST_VIRGINIA.txt",sep=" ")
names(Data)<-c("State","Year","Month", "Pop21", "Beer", "Spir", "Wine")
# CALCULATE CONSUMPTION PER 100,000 people 18 years and over
BeerHead=10*Data$Beer/Data$Pop21;

BeerHeadts<-ts(BeerHead,start=Data$Year[1],frequency=12)

# Figure 6.2:
#pdf("WestVaBeerHead.pdf")
par(mfrow=c(3,1))
ts.plot(BeerHeadts )
acf(BeerHeadts )
pacf(BeerHeadts)

#dev.off()
```


```{r, message=FALSE}
# Seasonally differenced data:
BeerHead.d12<-diff(BeerHeadts,lag=12)
# Figure 6.3:
#pdf("WestVaBeerHead_d12.pdf")
par(mfrow=c(3,1))
ts.plot(BeerHead.d12)
abline(h=0)
acf(BeerHead.d12)
pacf(BeerHead.d12)
#dev.off()
```


```{r, message=FALSE}
# Differenced and Seasonally Differenced Data
BeerHead.d12d1<-diff(diff(BeerHeadts,lag=12))
# Figure 6.4:
#pdf("WestVaBeerHead_d12d1.pdf")
par(mfrow=c(3,1))
ts.plot(BeerHead.d12d1)
abline(h=0)
abline(v=c(1981+7/12,1990+3/12),col=c("red","green"))

acf(BeerHead.d12d1)
pacf(BeerHead.d12d1)
#dev.off()
```


```{r, message=FALSE}

library(astsa)
#pdf("WestVaBeerHead_SARIMA_000_011.pdf")
sarima000011cons<-sarima(BeerHeadts,0,0,0,0,1,1,12)
pacf(sarima000011cons$fit$residuals)
#dev.off()

sarima000011cons    # Model results

shapiro.test(sarima000011cons$fit$residuals)

#pdf("WestVaBeerHead_SARIMA_011_011.pdf")

sarima011011<-sarima(BeerHeadts,0,1,1,0,1,1,12)

sarima011011   # Model results

#dev.off()
shapiro.test(sarima011011$fit$residuals)

Box.test(sarima011011$fit$residuals,lag=12,fitdf=2,type="Ljung-Box")
```



##########################################################################################

## Suggestions for Data Analysis

##########################################################################################

**NOTATION**: I USE the model name part 'sarima' for models fit using the 'sarima' command
 and 'sarma' for models fit below using the 'arima' command fit to differenced data.

### Fitting seasonal arima models using the built in R functions

Fitting ARIMA models to Seasonally Differenced Data
using original series, but specifying a seasonal lag difference in model:

```{r}
BeerHead.d12.sarma.alt<-arima(BeerHeadts,order=c(0,0,0),
                             seasonal=list(order=c(0,1,1),period=12) ,include.mean=TRUE) 
BeerHead.d12.sarma.alt    # gives results
```

#########################################################################################
**NOTE**: *include.mean=TRUE* is ignored and the following gives same result:
```{r}
BeerHead.d12.sarma.alt<-arima(BeerHead.d12,order=c(0,0,0),
                             seasonal=list(order=c(0,0,1),period=12),include.mean=FALSE )
```


Now using the seasonally differenced data but with mean term included:

```{r}
BeerHead.d12.sarma.cons<-arima(BeerHead.d12,order=c(0,0,0),
                               seasonal=list(order=c(0,0,1),period=12),include.mean=TRUE )
BeerHead.d12.sarma.cons
```

**compare with 'sarima000011cons' model above.**

But, *sarima000011cons* and *BeerHead.d12.sarima.alt* differ in at least one paremeter.
Look at the *sarima* function code to explain the difference.

### Fitting ARIMA models to Differenced and Seasonally Differenced Data
```{r}
BeerHead.d12d1.sarma<-arima(BeerHeadts,order=c(0,1,1),
                             seasonal=list(order=c(0,1,1),period=12))
BeerHead.d12d1.sarma  # model results

```


- Compare with *sarima011011* model above - are there any differences?

- Explain why *sarima011011* does not include an estimate of the constant term by 
 looking carefully at the 'sarima' function code.

### Fit model to differenced and seasonally differenced data including a constant: 
```{r}
BeerHead.d12d1.sarma.alt.cons<-arima(BeerHead.d12d1,order=c(0,0,1),
                                      seasonal=list(order=c(0,0,1),period=12),include.mean=TRUE )
BeerHead.d12d1.sarma.alt.cons
```


**NOTE**: intercept (contstant) not significant....Can you explain this?

###############################################################################################################
### Improved Model?

```{r}
library(astsa)

# My best model
sarima113011=sarima(BeerHeadts,1,1,3,0,1,1,12)
sarima113011

# MA(3) needed!


# Students attempts 2017:

sarima212011.KL=arima(BeerHeadts,order=c(2,1,2),seasonal=list(order=c(0,1,1),period=12))

sarima212011.KL


sarima1711011.AD<-arima(BeerHeadts,order=c(17,1,1),seasonal=list(order=c(0,1,1),period=12),
                    fixed = c(rep(0,13),rep(NA,1),rep(0,1),rep(NA,4)))
sarima1711011.AD

sarima1611011.SL<-arima(BeerHeadts,order=c(16,1,1),seasonal=list(order=c(0,1,1),period=12),
                     fixed = c(rep(0,13),rep(NA,1),rep(0,1),rep(NA,3)))
sarima1611011.SL

sarima016011.BD=arima(BeerHeadts,order=c(0,1,6), 
                   seasonal=list(order=c(0,1,1),period=12), fixed = c(NA,0,0,0,0,NA,NA))
sarima016011.BD


# library(fGarch)

# garchBeer=garchFit(sarima113011$residuals~garch(1,1),cond.dist="std")
# plot(garchBeer)

# ts.plot(BeerHead.d12d1)
# garchBeerd12d1=garchFit(BeerHead.d12d1~arma(1,0)+garch(1,1),cond.dist="std")
# 
# GARCH to differenced and seasonally differenced data directly.
# library(fGarch)
# garchBeerd12d1=garchFit(BeerHead.d12d1~garch(1,1),cond.dist="std")
# acf(garchBeerd12d1@residuals/garchBeerd12d1@sigma.t)
# Box.test(garchBeerd12d1@residuals/garchBeerd12d1@sigma.t,lag=24)

```

