---
title: "MATH5845 --- Time Series Analysis: Week 5"
author:
- Atefeh Zamani
- Deparment of Statistics
- University of New South Wales
outpute:
  pdf_document: default
  html_notebook: default
output:
  html_document:
    df_print: paged
---


# Packages

```{r, message=FALSE}
library(astsa)
#data(package="astsa")
library(itsmr)
```


### Simulated Data from ARIMA(l, 1, 0) Model (Example 6.3)

Simulate 200 observation from ARIMA(l, 1, 0) with $\phi = 0.8$ and $\sigma_W = 1$. Plot its sample autocorrelation and partial autocorrelation functions.
```{r}
set.seed(2023)
sim_data=astsa::sarima.sim(ar = 0.8, d = 1, ma =0, n = 200, t0 =0)

par(mfrow=c(3,1))
plot.ts(sim_data)
acf(as.numeric(sim_data),lag.max=40)
pacf(as.numeric(sim_data),lag.max=40)

sim_data_diff=diff(sim_data)
par(mfrow=c(3,1))
plot.ts(sim_data_diff)
acf(as.numeric(sim_data_diff),lag.max=40)
pacf(as.numeric(sim_data_diff),lag.max=40)

sim_estimate_mle1=ar.mle(sim_data_diff,aic = TRUE, order.max = 1)
sim_estimate_mle1

sim_estimate_ols2=ar.ols(sim_data, aic = TRUE, order.max = 2)
sim_estimate_ols2

```

### IMA(1, 1) and EWMA (Example 6.5)

Simulate 100 observations from an IMA(1,1)model with $\lambda= - \theta = 0.8$ and then calculate and
display the fitted EWMA superimposed on the data. This is accomplished using the
${\tt Holt-Winters}$ command in ${\tt R}.

```{r}
set.seed(2023)    
x = sarima.sim(ma = -0.8, d=1, n = 100)
plot.ts(x)
acf2(as.numeric(x),max.lag=40)

(x.ima = HoltWinters(x, beta=FALSE, gamma=FALSE))  # alpha is 1-lambda.
# beta parameter of Holt-Winters Filter. If set to FALSE, the function will do exponential smoothing.
plot(x.ima)

```


### Analysis of GNP Data (Example 6.6)

Consider the analysis of quarterly U.S. GNP from 1947(1) to 2002(3), $n = 223$ observations. The data are real U.S. gross national product in billions of chained 1996 dollars and have been seasonally adjusted.

Plot the data along with its ACF and PACF.
```{r}
par(mfrow=c(3,1))
gnp=ts(gnp, start=1947, frequency =4)
tsplot(gnp[1:100], col=4)
acf(gnp, 50)          
pacf(gnp, 50)
```

Calculate the growth rate and plot its time series along with the sample ACF and PACF. 
```{r}
gnpgr = diff(log(gnp))      # growth rate:diff(log(x))

#dev.new()
tsplot(gnpgr, col=4)
abline(h=mean(gnpgr))

#dev.new()
acf2(gnpgr, 24)  
```

Suggest some models for the data and check their goodness of fit. 
```{r}
sarima(gnpgr, 1, 0, 0)      # AR(1)
sarima(gnpgr, 0, 0, 2)      # MA(2) 
sarima(gnpgr, 0, 0, 2, no.constant=TRUE) # MA(2) without drift
ARMAtoMA(ar=.35, ma=0, 10)  # prints psi-weights


library(forecast)
Arima(gnpgr, c(1, 0, 0))
Arima(gnpgr, c(0, 0, 2))


auto.arima(gnpgr )

auto.arima(gnp,lambda="auto")
```


### A Seasonal AR Series (Example 6.7) 

Simulate a monthly data for 3 years which follows a first-order seasonal autoregressive series with $\Phi=0.9$. 
```{r}
set.seed(666)
SAR = sarima.sim(sar=.9, S=12, n=37) + 50
layout(matrix(c(1,2, 1,3), nc=2), heights=c(1.5,1))
tsplot(SAR, type="c", xlab="Year")
 abline(v=1:3, col=4, lty=2)
 points(SAR, pch=Months, cex=1.35, font=4, col=1:6)

phi  = c(rep(0,11),.9)
ACF  = ARMAacf(ar=phi, ma=0, 100)[-1] # [-1] removes 0 lag
PACF = ARMAacf(ar=phi, ma=0, 100, pacf=TRUE)
 LAG = 1:100/12   # time index
 # LAG1=1:100
tsplot(LAG, ACF, type="h", xlab="LAG", ylim=c(-.04,1))
# tsplot(LAG1, ACF, type="h", xlab="LAG", ylim=c(-.04,1))
 abline(h=0, col=8)
tsplot(LAG, PACF, type="h", xlab="LAG", ylim=c(-.04,1))
# tsplot(LAG1, PACF, type="h", xlab="LAG", ylim=c(-.04,1))
 abline(h=0, col=8)
 
# acf(SAR)
# pacf(SAR)
```

### A Mixed Seasonal Model (Example 6.8)

Create the theoretical ACF and PACF of ARMA$(0, 1) \times  (1, 0)_{12}$ with with $\Phi$ = .8 and $\theta$ = -.5.
```{r}
par(mfrow=1:2)
phi  = c(rep(0,11),.8)
ACF  = ARMAacf(ar=phi, ma=-.5, 50)[-1]
PACF = ARMAacf(ar=phi, ma=-.5, 50, pacf=TRUE)
 LAG = 1:50/12
tsplot(LAG, ACF, type="h", xlab="LAG", ylim=c(-.4,.8), col=4, lwd=2)
 abline(h=0, col=8)
tsplot(LAG, PACF, type="h", xlab="LAG", ylim=c(-.4,.8), col=4, lwd=2)
 abline(h=0, col=8)
```


### Example 5.14: Air Passengers
```{r}
x     = AirPassengers
log_x    = log(x) 
dlog_x   = diff(log_x) 
ddlog_x  = diff(dlog_x, 12)
tsplot(cbind(x, log_x, dlog_x, ddlog_x), main="")


acf2(as.numeric(ddlog_x,50))
# below of interest for showing seasonal persistence (not shown here):
par(mfrow=c(2,1))
monthplot(dlog_x)
monthplot(ddlog_x)

sarima(log_x, 1,1,1, 0,1,1, 12)   # model 1
sarima(log_x, 0,1,1, 0,1,1, 12)   # model 2 (the winner)
sarima(log_x, 1,1,0, 0,1,1, 12)   # model 3

dev.new()
sarima.for(log_x, 12, 0,1,1, 0,1,1,12)  # forecasts



library(forecast)
auto.arima(x, lambda="auto", seasonal = TRUE)

```

