---
title: 'Chapter 2: Analyse Lake Huron Levels'
output:
  pdf_document: default
  html_document: default
  word_document: default
--- 

Read and plot series

```{r}
library(astsa)
LakeLevel<-scan('LakeHuronLevels.txt')
ts.plot(ts(LakeLevel,start=1875),ylab="Level (ft)")
```

Fit time trend line using least squares regression on time trend
Some possible curvature in trend on time so try quadratic in time:

```{r}

Time<-1:98

LSfit<-lm(LakeLevel~Time)

summary(LSfit)


LSfit2<-lm(LakeLevel~Time+I(Time^2))
plot(Time,LakeLevel,type="l")
lines(LSfit$fitted,col="red")
lines(LSfit2$fitted,col="green")
summary(LSfit2)
```

Note that the quadratic term is significant.
Compare autocorrelation functions of residuals from linear and quadratic trend models:

```{r}
par(mfrow=c(2,1))
acf2(LSfit$residuals)
acf2(LSfit2$residuals)
```

For both the linear and quadratic trends the residual ACF is like an AR(1).
Fit an AR(1)  model using maximum likelihood and calculate the significance 
of the linear trend.

```{r}
LakeLevels.AR1<-arima(LakeLevel,order=c(1,0,0),xreg=Time)
LakeLevels.AR1

Trendzval<-LakeLevels.AR1$coef[3]/diag(LakeLevels.AR1$var.coef)[3]^0.5
Trendpval<-2*pnorm(-abs(Trendzval))
print(Trendpval)
```

Define a function to summarise the model in standard table form, 
apply to the linear regression plus AR(1) fit.

```{r}
arimamodelsummary<-function(arimamodel){
  coef<-arimamodel$coef
  se<-diag(arimamodel$var.coef)^0.5
  zstat<-coef/se
  pval<-2*pnorm(-abs(zstat))
  results<-cbind(coef,se,zstat,pval)
  colnames(results)<-c("Estimate","s.e.","z-val","p-val")
  print(round(results,4))
}
arimamodelsummary(LakeLevels.AR1)
```

Check if residuals are now showing properties of Gaussian white noise.

```{r}
par(mfrow=c(1,2))
acf(LakeLevels.AR1$res)
qqnorm(LakeLevels.AR1$res)
```

Still some residual autocorrelation at lag 1 so try AR(2) for residuals:

```{r}
LakeLevels.AR2<-arima(LakeLevel,order=c(2,0,0),xreg=Time)
arimamodelsummary(LakeLevels.AR2)
par(mfrow=c(1,2))
acf(LakeLevels.AR2$res)
qqnorm(LakeLevels.AR2$res)
```


How about an ARMA(1,1)....?:

```{r}
LakeLevels.ARMA11<-arima(LakeLevel,order=c(1,0,1),xreg=Time)
LakeLevels.ARMA11
arimamodelsummary(LakeLevels.ARMA11)

par(mfrow=c(1,2))
acf(LakeLevels.ARMA11$res)
qqnorm(LakeLevels.ARMA11$res)
```

Compare the fit of  AR(1), AR(2) and ARMA(1,1) to the data:

```{r}
par(mfrow=c(1,1))
plot(Time,LakeLevel,type="l")
lines(Time,LakeLevel-LakeLevels.AR1$residuals,col="blue")
lines(Time,LakeLevel-LakeLevels.AR2$residuals,col="red")
lines(Time,LakeLevel-LakeLevels.ARMA11$residuals,col="green")
```

Discuss and conclude....

Is the quadratic trend significant??

```{r}
Time2<-Time^2
LakeLevels.QuadTrend.AR2<-arima(LakeLevel,order=c(2,0,0),xreg=cbind(Time,Time2))
arimamodelsummary(LakeLevels.QuadTrend.AR2)
par(mfrow=c(1,2))
acf(LakeLevels.QuadTrend.AR2$res)
qqnorm(LakeLevels.QuadTrend.AR2$res)
```
