---
title: 'Tutorial: Week 2- Solution'
author:
  - Deparment of Statistics
  - University of New South Wales
output:
  pdf_document: default
  html_notebook: default
  word_document: default
  output: pdf_document
---
  
### Packages
  
```{r, message=FALSE}
library(astsa)
#data(package="astsa")
library(itsmr)
library("dygraphs")
```

### Dow Jones Utilities Index

Using the data in the file *DowJonesUtil.txt*

- replicate the analysis and conclusions in Section 3.1.5. 

```{r}
# Convert the 'dowj' data (from itsmr package) into a time series object
DJUtilIndex = ts(dowj)

# Set up the graphics window to show 2 rows and 2 columns of plots
par(mfrow=c(2,2))

# Plot the original Dow Jones Utilities Index time series
ts.plot(DJUtilIndex, main = "Original Series")

# Plot the autocorrelation function (ACF) of the original series
acf(DJUtilIndex, main = "ACF of Original Series")

# Take the first difference to remove trend and make the series stationary
DJUtilIndex.D1 <- diff(DJUtilIndex)

# Plot the differenced series
ts.plot(DJUtilIndex.D1, main = "Differenced Series (Lag 1)")
abline(h=0)  # Add a horizontal reference line at 0

# Plot the ACF of the differenced series
acf(DJUtilIndex.D1, main = "ACF of Differenced Series")

DJUtilIndex.arima<- arima(DJUtilIndex, order = c(1,1,1), 
                                include.mean = FALSE, method = "CSS")
DJUtilIndex.arima

# Fit an ARMA(1,1) model to the differenced data using Conditional Sum of Squares (CSS) method
DJUtilIndex.ARMA11_CSS <- arima(DJUtilIndex.D1, order = c(1,0,1), 
                                include.mean = FALSE, method = "CSS")
DJUtilIndex.ARMA11_CSS
# Fit an ARMA(1,1) model to the differenced data using Maximum Likelihood (ML) method
DJUtilIndex.ARMA11_ML <- arima(DJUtilIndex.D1, order = c(1,0,1), 
                               include.mean = FALSE, method = "ML")

# Display the result of the ML-fitted ARMA(1,1) model
DJUtilIndex.ARMA11_ML

```


- Check the assumptions that the residuals from the ARMA(1,1) model are normal and uncorrelated. 

```{r}
res_dj = DJUtilIndex.ARMA11_ML$residuals

plot(res_dj, main = "Residuals from the ARMA(1,1) model")
```
To check normaility, you can use histogram, qqplot and shapiro test:

```{r}
# Histogram
hist(res_dj, breaks = 10, probability = TRUE, xlim = c(-1.5,2), xlab = "Residuals", main = "Histogram of residuals from the ARMA(1,1) model")
curve(dnorm(x, mean=mean(res_dj), sd=sd(res_dj)), add=TRUE, col="red")

# Q-Q plot
qqnorm(res_dj)
qqline(res_dj, col = "red")

# Shapiro-Wilk test
shapiro.test(res_dj)
# Null hypothesis: data are normally distributed.
# Small p-value --> reject normality.
```
Uncorrelatedness can be checked visually using ACF plot of residuals or we can use Ljung-Box test
```{r}
# ACF plot of residuals
acf(res_dj, main = "ACF of residuals from the ARMA(1,1) model")

#Ljung-Box test (for white noise)
Box.test(res_dj, lag = 20, type = "Ljung-Box")
#Null hypothesis: residuals are uncorrelated (white noise).
#Small p-value --> autocorrelation exists --> model may be inadequate.
```


- Fit the AR(1), the MA(1) and the ARMA(1,1). Compare your results with those reported in Section 3.1.5.

- Plot the observed ACF and the model ACF (theoretical using the model estimates) for the three models considered. Discuss.

```{r}
# Fit an AR(1) model to the differenced data (no intercept)
DJUtilIndex.AR1 <- arima(DJUtilIndex.D1, order = c(1, 0, 0), include.mean = FALSE)

# Display the AR(1) model summary
DJUtilIndex.AR1

# Fit a MA(1) model to the differenced data (no intercept)
DJUtilIndex.MA1 <- arima(DJUtilIndex.D1, order = c(0, 0, 1), include.mean = FALSE)

# Display the MA(1) model summary
DJUtilIndex.MA1

# Set up the graphics window to show 2 rows and 2 columns of plots
par(mfrow = c(2, 2))

# Plot the observed ACF of the differenced data
acf(DJUtilIndex.D1, lag.max = 15, main = "Observed ACF DJ Utilities Index")

# Compute and plot the theoretical ACF for the fitted AR(1) model
AR1acf <- ARMAacf(ar = coef(DJUtilIndex.AR1), ma = 0, lag.max = 15)
plot(as.integer(names(AR1acf)), AR1acf, type = "h", xlab = "lag", ylim = c(-0.2, 1))
abline(h = 0)
title("Theoretical ACF AR(1) Fit")

# Compute and plot the theoretical ACF for the fitted MA(1) model
MA1acf <- ARMAacf(ar = 0, ma = coef(DJUtilIndex.MA1), lag.max = 15)
plot(as.integer(names(MA1acf)), MA1acf, type = "h", xlab = "lag", ylim = c(-0.2, 1))
abline(h = 0)
title("Theoretical ACF MA(1) Fit")

# Compute and plot the theoretical ACF for the fitted ARMA(1,1) model
ARMA11acf <- ARMAacf(
  ar = coef(DJUtilIndex.ARMA11_ML)[1],
  ma = coef(DJUtilIndex.ARMA11_ML)[2],
  lag.max = 15
)
plot(as.integer(names(ARMA11acf)), ARMA11acf, type = "h", xlab = "lag", ylim = c(-0.2, 1))
abline(h = 0)
title("Theoretical ACF ARMA(1,1) Fit")

```
    
### Overshorts analysis

Consider the overshorts data in *overshorts.txt*. 

- Perform a standard t-test (use the *R* command *t.test(..)*) and draw your conclusions about the hypothesis that the mean level of overshorts is zero versus the alternative that the tank is not leaking (one-sided alternative since we really do not expect the tank to be gaining, unless the pump meter is dodgy!). 

The analysis summarized here is based on Brockwell and Davis (2002). The data were collected as part of a large scale routine monitoring of gasoline stations in Colorado.

-  $X_{t}:$ the amount measured using a dipstick in the inground petrol storage tank at the end of day $t$

- $A_{t}:$ the amount of petrol sold during the day minus any amount that was delivered during the day. 

We expect that 
\[
    X_{t} \approx X_{t-1} + {\tt delivered}_t - {\tt sold}_t =  X_{t-1} -A_t.
\]
Let the `overshorts' be denoted by

\[
Y_{t}=X_{t}-X_{t-1}+A_{t}%
\]

Under the assumption of no measurement error and no leakage in the tank we should have $Y_{t}\equiv0$. In practice there is measurement variability. In this case the `overshorts' $Y_{t}$ will be samples from a distribution with mean $\mu$. If the tank is not leaking into the subsoil then we should have $H_{0}:\mu=0.$ We want to test this null hypothesis against the alternative that the tank is leaking $H_{a}:\mu<0.$
```{r}
overshorts=scan('overshorts.txt')
ts.plot(overshorts)
abline(h=0)

#t_test_stat=(mean(overshorts))/(sqrt(var(overshorts)/length(overshorts)))

t.test(overshorts, alternative = "less")
```

- Look at the sample ACF for the overshorts series, discuss if the MA(1) model seems appropriate, fit this model using the *arima*. Command and reconsider your previous conclusion concerning the hypothesis that the tank is not leaking. 

```{r}
acf(overshorts)
overshorts.ma1<-arima(overshorts,order=c(0,0,1), include.mean=T, method= "ML")
overshorts.ma1
t_stat=overshorts.ma1$coef[2]/sqrt(overshorts.ma1$var.coef[2,2])
t_stat
pt(t_stat, df=56)
```


- Check the properties of the residuals. 
```{r}
model_Res = overshorts.ma1$residuals
plot(model_Res, main = "Residuals from fitting MA(1) to overshorts data")

# Normality
# Q-Q plot
qqnorm(model_Res)
qqline(model_Res, col = "red")

# Shapiro-Wilk test
shapiro.test(model_Res)

# Uncorrelatedness
acf(model_Res, main = "ACF of residuals from fitting MA(1) to overshorts data")
Box.test(model_Res, lag = 20, type = "Ljung-Box")
```

- Plot the model ACF for the fitted model and compare with the observed ACF.
```{r}
par(mfrow = c(2,2))
acf(overshorts)
pacf(overshorts)

MA1acf <- ARMAacf(ar = 0, ma = coef(overshorts.ma1), lag.max = 15)
plot(as.integer(names(MA1acf)), MA1acf, type = "h", xlab = "lag", ylim = c(-0.2, 1))
abline(h = 0)
title("Theoretical ACF MA(1) Fit")


MA1pacf <- ARMAacf(ar = 0, ma = coef(overshorts.ma1), lag.max = 15, pacf = TRUE)
plot( MA1pacf, type = "h", xlab = "lag", ylim = c(-0.2, 1))
abline(h = 0)
title("Theoretical PACF MA(1) Fit")
```


### Lake Huron analysis

The data are in *LakeHuronLevels.txt* and an *R*-script which performs most of the analysis required is in *LakeHuronAnalysis.Rmd*. 

- Use this file to understand how to assess serial dependence in simple regression settings and to include modelling serial dependence in the regression modelling. 

- Experiment with alternative specifications of the autocorrelation in the residuals and record your conclusions.

*Solution*
Read and plot series

```{r}
library(astsa)
LakeLevel<-scan('LakeHuronLevels.txt')
ts.plot(ts(LakeLevel,start=1875),ylab="Level (ft)")
```

Fit time trend line using least squares regression on time trend
Some possible curvature in trend on time so try quadratic in time:

```{r}

Time<-1:98

LSfit<-lm(LakeLevel~Time)

summary(LSfit)


LSfit2<-lm(LakeLevel~Time+I(Time^2))
plot(Time,LakeLevel,type="l")
lines(LSfit$fitted,col="red")
lines(LSfit2$fitted,col="green")
summary(LSfit2)
```

Note that the quadratic term is significant.
Compare autocorrelation functions of residuals from linear and quadratic trend models:

```{r}
par(mfrow=c(2,1))
acf2(LSfit$residuals)
acf2(LSfit2$residuals)
```

For both the linear and quadratic trends the residual ACF is like an AR(1).
Fit an AR(1)  model using maximum likelihood and calculate the significance 
of the linear trend.

```{r}
LakeLevels.AR1<-arima(LakeLevel,order=c(1,0,0),xreg=Time)
LakeLevels.AR1

Trendzval<-LakeLevels.AR1$coef[3]/diag(LakeLevels.AR1$var.coef)[3]^0.5
Trendpval<-2*pnorm(-abs(Trendzval))
print(Trendpval)
```

Define a function to summarise the model in standard table form, 
apply to the linear regression plus AR(1) fit.

```{r}
arimamodelsummary<-function(arimamodel){
  coef<-arimamodel$coef
  se<-diag(arimamodel$var.coef)^0.5
  zstat<-coef/se
  pval<-2*pnorm(-abs(zstat))
  results<-cbind(coef,se,zstat,pval)
  colnames(results)<-c("Estimate","s.e.","z-val","p-val")
  print(round(results,4))
}
arimamodelsummary(LakeLevels.AR1)
```

Check if residuals are now showing properties of Gaussian white noise.

```{r}
par(mfrow=c(1,2))
acf(LakeLevels.AR1$res)
qqnorm(LakeLevels.AR1$res)
qqline(LakeLevels.AR1$res, col = "red")
```

Still some residual autocorrelation at lag 1 so try AR(2) for residuals:

```{r}
LakeLevels.AR2<-arima(LakeLevel,order=c(2,0,0),xreg=Time)
LakeLevels.AR2$aic

arimamodelsummary(LakeLevels.AR2)
par(mfrow=c(1,2))
acf(LakeLevels.AR2$res)
qqnorm(LakeLevels.AR2$res)
qqline(LakeLevels.AR2$res, col = "red")
```


How about an ARMA(1,1)....?:

```{r}
LakeLevels.ARMA11<-arima(LakeLevel,order=c(1,0,1),xreg=Time)
LakeLevels.ARMA11
arimamodelsummary(LakeLevels.ARMA11)

par(mfrow=c(1,2))
acf(LakeLevels.ARMA11$res)
qqnorm(LakeLevels.ARMA11$res)
qqline(LakeLevels.ARMA11$res, col = "red")

```

Compare the fit of  AR(1), AR(2) and ARMA(1,1) to the data:

```{r}
par(mfrow=c(1,1))
plot(Time,LakeLevel,type="l")
lines(Time,LakeLevel-LakeLevels.AR1$residuals,col="blue")
lines(Time,LakeLevel-LakeLevels.AR2$residuals,col="red")
lines(Time,LakeLevel-LakeLevels.ARMA11$residuals,col="green")
```

Discuss and conclude....

Is the quadratic trend significant??

```{r}
Time2<-Time^2
LakeLevels.QuadTrend.AR2<-arima(LakeLevel,order=c(2,0,0),xreg=cbind(Time,Time2))
LakeLevels.QuadTrend.AR2$aic

arimamodelsummary(LakeLevels.QuadTrend.AR2)
par(mfrow=c(1,2))
acf(LakeLevels.QuadTrend.AR2$res)
qqnorm(LakeLevels.QuadTrend.AR2$res)
qqline(LakeLevels.QuadTrend.AR2$res, col = "red")

```
