---
title: "Data 9001: Tutorial Week 1"
author: "Fangzhou Yu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{=html}
<!--
DATA 9001
UNSW 2024 T2
Author: Martin Eftimoski
This version: Fangzhou Yu 2025 T2
-->
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

# Exercise 1: Basic R Operations

**Question:** Access R through your preferred method (R Studio/Posit). And run the following code:

```{r}
# Variables
x <- 5
y <- 3
z <- x + y

# Functions
z <- sqrt(z)
```

**What is the value of x, y, and z?**

```{r, echo=FALSE}
# Display the values
cat("x =", x, "\n")
cat("y =", y, "\n") 
cat("z =", z, "\n")
```

**Answer:** - X = 5 - Y = 3\
- Z = 2.828427

Note that R operates via the assign operator " \<- ". In most of the cases, it is equivalent to the equals sign " = ".

# Exercise 2: Data Loading and Variable Definition

**Question:** Load the data "Spend_data.csv" and "Store_data.csv" by reading the csv files into the environment. Construct a table defining all the variables in the dataset for each of the two datasets. Define the data type for each of the variables.

```{r}
# Load the datasets
Spend_data <- read.csv("Advertising_spend.csv")
Store_data <- read.csv("Store_information.csv")

# Display structure of datasets
head(Spend_data)
head(Store_data)
```

**Variable Definitions:**

| Dataset    | Variable          | Data Type                 |
|------------|-------------------|---------------------------|
| Spend_data | Store_ID          | Categorical (Nominal)     |
| Spend_data | Sales             | Quantitative (Continuous) |
| Spend_data | Advertising_Spend | Quantitative (Continuous) |
| Spend_data | Season            | Categorical (Nominal)     |
| Store_data | Store_ID          | Categorical (Nominal)     |
| Store_data | Region            | Categorical (Nominal)     |
| Store_data | Product_Category  | Categorical (Nominal)     |
| Store_data | Customer_Rating   | Categorical (Ordinal)     |

# Exercise 3: Package Installation and Data Merging

**Question:** Install and load the following packages: - ggplot2 – package used for visualisation - dplyr – package used for data manipulation/cleaning\
- tidyr – package used for data transformation

Merge the two datasets using the variable "Store_ID". Left join the table "Store_data" onto the table "Spend_data". Visually inspect the data. What do you notice?

```{r}
# Install and load required packages
# install.packages(c("ggplot2", "dplyr", "tidyr"))
library(ggplot2)
library(dplyr)
library(tidyr)

# Alternatively, you can load the tidyverse package which includes ggplot2, dplyr, and tidyr by
# install.packages("tidyverse")
# library(tidyverse)

# Merge the datasets
merged_data <- merge(Spend_data, Store_data, by = "Store_ID", all = TRUE)

# Display first few rows
head(merged_data)
```

**Observation:** The data has missing observations as seen by `<NA>`.

# Exercise 4: Data Cleaning and Missing Value Handling

**Question:** Process the merged dataset by handling any missing data appropriately. In the case of quantitative variables this can be via replacing the missing data with the mean. In the case of ordinal categorical variables this could be using the median. Or in the case of other variables it must simply be left "unknown". Export the data to a csv file, "merged_data.csv".

```{r}
# Cleaning: Handling missing values

# Fill missing 'Sales' and 'Advertising_Spend' with mean
merged_data$Sales[is.na(merged_data$Sales)] <- mean(merged_data$Sales, na.rm = TRUE)
merged_data$Advertising_Spend[is.na(merged_data$Advertising_Spend)] <- mean(merged_data$Advertising_Spend, na.rm = TRUE)

# Fill missing 'Season' with 'Unknown'
merged_data$Season[is.na(merged_data$Season)] <- "Unknown"

# Fill missing 'Region' and 'Product_Category' with 'Unknown'
merged_data$Region[is.na(merged_data$Region)] <- "Unknown"
merged_data$Product_Category[is.na(merged_data$Product_Category)] <- "Unknown"

# Fill missing 'Customer_Rating' with median
merged_data$Customer_Rating[is.na(merged_data$Customer_Rating)] <- median(merged_data$Customer_Rating, na.rm = TRUE)

# Export the cleaned and merged dataset to CSV
write.csv(merged_data, file = "merged_cleaned_data.csv", row.names = FALSE)

# Display the first few rows of the merged and cleaned dataset
head(merged_data)
```

## Alternative Solution: Using dplyr and tidyr

As discussed earlier, we can also use the tidyverse approach with `replace_na()` from tidyr, which is more readable and pipe-friendly:

```{r}
# Alternative solution: use replace_na() from tidyr
merged_data <- merged_data %>%
  mutate(
    Sales = replace_na(Sales, mean(Sales, na.rm = TRUE)),
    Advertising_Spend = replace_na(Advertising_Spend, mean(Advertising_Spend, na.rm = TRUE)),
    Season = replace_na(Season, "Unknown"),
    Region = replace_na(Region, "Unknown"),
    Product_Category = replace_na(Product_Category, "Unknown"),
    Customer_Rating = replace_na(Customer_Rating, median(Customer_Rating, na.rm = TRUE))
  )
```

# Exercise 5: Basic Scatter Plot Analysis

**Question:** Provide an appropriate graph to investigate if there is a relationship between advertising spend and sales across all stores. Interpret the graph.

```{r}
# Scatter Plot of Sales vs. Advertising Spend
ggplot(merged_data, aes(x = Advertising_Spend, y = Sales)) +
  geom_point(color = "blue", size = 2) +
  labs(title = "Sales vs Advertising Spend",
       x = "Advertising Spend ($)",
       y = "Sales ($)") +
  theme_minimal()
```

**Interpretation:** There seems to be a relationship between increasing advertising spend and sales. But it is very noisy.

# Exercise 6: Adding Color Dimension

**Question:** Using colour as a 3rd dimension for the data, investigate if there is a relationship between advertising spend and sales that varies with seasons across all stores. How can this graph be improved?

```{r}
# Define a color palette for seasons
season_colors <- c(
  "Winter" = "#1f78b4",  # Blue
  "Spring" = "#33a02c",  # Green
  "Summer" = "#e31a1c",  # Red
  "Autumn" = "#ff7f00"   # Orange
)

# Scatter Plot of Sales vs. Advertising Spend by Season
ggplot(merged_data, aes(x = Advertising_Spend, y = Sales, color = Season)) +
  geom_point(size = 2) +
  scale_color_manual(values = season_colors) +
  labs(title = "Sales vs Advertising Spend by Season",
       x = "Advertising Spend ($)",
       y = "Sales ($)") +
  theme_minimal()
```

**Interpretation:** Despite adding colour to represent the seasons the data is too noisy for it to be clear on one graph. This reflects the weak seasonal relationship or the lack of data.

## Alternative Solution: Using Automatic Color Palette

Instead of manually defining colors, we can use automatic color palette generators like `scale_color_viridis_d()` which are colorblind-friendly and perceptually uniform:

```{r}
# Exercise 6 - Alternative using viridis palette
ggplot(merged_data, aes(x = Advertising_Spend, y = Sales, color = Season)) +
  geom_point(size = 2) +
  scale_color_viridis_d() +  # Automatic viridis discrete color palette
  labs(title = "Sales vs Advertising Spend by Season (Viridis Palette)",
       x = "Advertising Spend ($)",
       y = "Sales ($)") +
  theme_minimal()
```

**Note:** The color palette was defined multiple times in the original code, which is unnecessary. It's more efficient to define it once at the beginning and reuse it throughout the analysis.

# Exercise 7: Faceted Scatter Plots

**Question:** Using the "facet_wrap" function of R, create 4 scatterplots for advertising spend and sales across all seasons. Interpret the results.

```{r}
# Facet Grid of Sales vs. Advertising Spend by Season
# Define a color palette for seasons
season_colors <- c(
  "Winter" = "#1f78b4",  # Blue
  "Spring" = "#33a02c",  # Green
  "Summer" = "#e31a1c",  # Red
  "Autumn" = "#ff7f00"   # Orange
)

ggplot(merged_data, aes(x = Advertising_Spend, y = Sales, color = Season)) +
  geom_point(size = 2) +
  facet_wrap(~ Season) +
  scale_color_manual(values = season_colors) +
  labs(title = "Sales vs Advertising Spend by Season",
       x = "Advertising Spend ($)",
       y = "Sales ($)") +
  theme_minimal()
```

**Interpretation:** There seems to be a relatively strong relationship between advertising spend and sales for Summer and Autumn. A weaker relationship for Winter. And seemingly no relationship for Spring.

# Exercise 8: Outlier Detection with Box Plots

**Question:** Identify if our relationship analysis is being influenced by any outliers in sales data. What graph would you use for this?

```{r}
# Boxplot of Sales by Season
# Define a color palette for seasons
season_colors <- c(
  "Winter" = "#1f78b4",  # Blue
  "Spring" = "#33a02c",  # Green
  "Summer" = "#e31a1c",  # Red
  "Autumn" = "#ff7f00"   # Orange
)

ggplot(merged_data, aes(x = Season, y = Sales, fill = Season)) +
  geom_boxplot() +
  scale_fill_manual(values = season_colors) +
  labs(title = "Distribution of Sales by Season",
       x = "Season",
       y = "Sales ($)") +
  theme_minimal()
```

**Interpretation:** A box plot is an effective way to quickly identify outliers. We can see that Summer has outliers that could be skewing the relationship as the sales data is less representative.

# Exercise 9: Distribution Analysis with Histograms

**Question:** Are there any systematic differences in advertising spend across seasons? What graph could help us investigate this?

```{r}
# Define a color palette for seasons
season_colors <- c(
  "Winter" = "#1f78b4",  # Blue
  "Spring" = "#33a02c",  # Green
  "Summer" = "#e31a1c",  # Red
  "Autumn" = "#ff7f00"   # Orange
)

# Histogram of Advertising Spend by Season
ggplot(merged_data, aes(x = Advertising_Spend, fill = Season)) +
  geom_histogram(binwidth = 1000, color = "black", alpha = 0.7, position = "identity") +
  facet_wrap(~ Season) +
  scale_fill_manual(values = season_colors) +
  labs(title = "Distribution of Advertising Spend by Season",
       x = "Advertising Spend ($)",
       y = "Count") +
  theme_minimal()
```

**Interpretation:** A histogram is a very useful tool to investigate the underlying distribution of data. In this case we can see that the empirical mean and variance will be different across seasons for advertising spend, also suggesting the relationship is further complicated.
